---
title: "Evaluate BP-SP model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{my-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This document illustrates how to estimate BP-SP models on a restricted
simulated dataset using `R` and `stan`.

## Dataset Overview

The analysis is based on a simulated dataset `dt_sim`, which consists
of gold-standard incidence $I$ observed in 183 postal codes
supposedly covered by cancer registries, and proxy counts $P$
observed in all of the 704 postal codes in the territory. The shape of
the analysed region is contained in the dataset `shap_geo`.  

The shape of the incidence rate, and the $se$ and $ppv$ values by age,
are taken from the estimates for Corpus Uteri cancer in women (CU)
with the hospitalisation proxy. The parameters used for the simulation
are the same as those used in the simulation section of the manuscript
for the $\rho=0.7$ and $N=20$ scenario. They are given in the
following table: 

```{r, include = TRUE, echo=F}
library(kableExtra)
library(tidyverse)
dt_par=tribble(~param,~value,
               "$N$",20,
               "$\\overline{Se}$",0.8,
               "$\\overline{PPV}$",0.7,
               "$\\sigma_c$",0.2,
               "$\\phi_c$",0.7,
               "$\\delta_1$",0.57,
               "$\\phi_1$",0.7,
               "$\\delta_2$",0.41,
               "$\\phi_2$",0.7,
               "$\\rho$",0.7
               )

dt_par %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Setup
The code used for the simulations is detailed in the file [simulate-data.html](simulate-data.html).

We first load the packages required for data manipulation, the
dataset, and some utility functions needed to format the datasets for
`stan` use: 

```{r setup}
library(tidyverse)
theme_set(theme_bw())
library(magrittr)
library(Matrix)
library(mgcv)
library(sf)
library(spdep)
library(RSpectra)

load(file="./data/shap_geo.rda") ## Shapefile
load(file="./data/dt_sim.rda") ## Simulated data

source("./R/format_data.R") ## R utilities to format data

```


Estimation of the model may be done with `stan` or `cmdstan`. We use
the latter in this vignette (see https://mc-stan.org/cmdstanr/ for
installation instructions), and we will use the `tidybayes` package
to format posterior estimates. 

```{r cmd}
library(cmdstanr)
library(tidybayes)
```

The first step is to compile the `stan` program implementing the BP-SP
model. This is done as follows: 

```{r}
bpsp <- cmdstan_model("./inst/stan/bp_sp.stan")

``` 

Next, the data must be processed to comply with the `stan` program
requirements. We built a dedicated function `prep_data` which
facilitates the data processing. 

The function has two data requirements:

1. `dt`: This must contain a `geo` column that identifies the
   geographical area, as well as the incidence and proxy counts, and
   any other covariate to be passed to the model. 
2. `geo`: This informs the geographical structure of the
   territory. The easiest format to use is an `sf` data.frame. 

And two formulas:

1. `form_id`: This specifies the within-area model. 
2. `form_geo`: This specifies the model for the spatial latent
   variables for incidence, $Se$, and $PPV$. Different specifications
   may be set for random effects (RE). For example, for an ICAR RE for
   $Se$, an IID for $PPV$, and BYM2 for incidence, the formula would
   be set as: 

   ```r
   form_geo = ~ i(bym(geo)) + se(icar(geo)) + ppv(iid(geo))
   ```


Here, we specify BYM2 prior for incidence, Se and PPV, and model age
via a cubic regression spline whith 5 knots, and put the log of
person-years as an offset for the component $\mu^I_{ij}$.

```{r}
stan_dt <- prep_data(dt = dt_sim,
                     form_id = c(I,P) ~ offset(log(py)) + s(age,bs="cr",k=5,fx=TRUE),
                     form_geo= ~ bym(geo),
                     geo=shap_geo)

``` 

The `stan_dt`object is a tibble with four components:

1. `stan_dt$dt_fit`: a list of data for the stan fit;
3. `stan_dt$dt_post`: a tibble used in post-processing stan fit;
4. `stan_dt$form_id`: recall of `form_id`
5. `stan_dt$form_geo`: recall of `form_geo`

## Stan fit

Now that the material is set, we can sample the HMC:

```{r,eval=F}
chains=4
iter=1000
warmup=1000
thin=1

fit <- bpsp$sample(stan_dt$dt_fit[[1]],
                   chains = chains, parallel_chains = chains,
                   iter_sampling = iter,iter_warmup = warmup,thin = thin,
                   seed = 1,
                   adapt_delta = 0.95,
                   show_exceptions = TRUE,
                   refresh = round(iter/10),
                   save_warmup =F)

``` 

The model took 10 minutes to run on a CentOS 7 Linux server with an
Intel(R) Xeon(R) Gold 2.90 GHz processor. The fit is made available in the
file [`fit.rda`](https://drive.google.com/file/d/1BQ39CyDaCrRTwoK_pDNTBecakQTvlsaf/view?usp=sharing).


```{r,eval=T}
library(googledrive)
drive_download(
  file="https://drive.google.com/file/d/1BQ39CyDaCrRTwoK_pDNTBecakQTvlsaf/view?usp=sharing", path = "./data/fit.rda", overwrite = TRUE)
load(file="./data/fit.rda")
```

## Posterior processing

The correspondence between the age and geo indexes in the Stan fit and
the original values is stored in the `stan_dt$dt_post` data.  
In the following, we build functions to establish this correspondence. 
```{r,eval=T}
  # Functions for posterior summary
  decod_age <- . %>%
    left_join(
      stan_dt$dt_post[[1]]$decod[[1]]$age,
      by=c("id_age")) %>%
    select(-id_age)
  
  decod_geo <- . %>%
    left_join(
      stan_dt$dt_post[[1]]$decod[[1]]$geo_ra %>%
        mutate(n_geo=n()) %>%
        bind_rows(
          stan_dt$dt_post[[1]]$decod[[1]]$geo %>%
            mutate(n_geo=n())),
      by=c("id_geo","n_geo")) %>%
    select(-id_geo)
  
```
  

### Parameters
The correspondence between the model parameters and their `stan` names
is given in the table below: 

```{r, include = TRUE, echo=F}
library(kableExtra)
dt_par=tribble(~param,~stan,
               "$\\overline{Se}$","b_sp[1]",
               "$\\overline{PPV}$","b_sp[2]",
               "$\\sigma_c$","s_c",
               "$\\phi_c$","phi_c",
               "$\\delta_1$","s_u[1]",
               "$\\phi_1$","phi_u[1]",
               "$\\delta_2$","s_u[2]",
               "$\\phi_2$","phi_u[1]",
               "$\\rho$","r_u[1]",
               "$\\sigma_{\\nu_1}$","s_nu[1]",
               "$\\sigma_{\\nu_2}$","s_nu[2]",
               "$\\kappa$","kappa",
               "$\\psi$","psi",
               "$\\sigma_e$","s_e",
               "$\\sigma_{d^\\star}$","s_ds"               
               )

dt_par %>%
  kbl() %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Their posterior summaries are recovered using:

```{r,eval=T}
  # Parameters
  par_sum <- fit %>% 
    gather_draws(s_c,s_u[comp],s_d,s_ds,s_e,kappa,psi,b_sp[comp],phi_c,phi_u[comp],r_u[comp],s_nu[comp]) %>% 
    tidybayes::summarise_draws("mean", "median","sd", "rhat","ess_bulk","ess_tail",
                               ~quantile(.x, probs = c(0.025, 0.975)))%>%
    rename(pm=mean,par=.variable,low=`2.5%`,up=`97.5%`) %>%
    ungroup() %>%
    select(-variable) %>%
    mutate(par=ifelse(is.na(comp),par,paste0(par,comp))) %>%
    select(-comp)
  
  par_sum
  
```

The convergence criteria `rhat` and `ess` are generally good for the
relatively small number of iterations used. The key parameters
$\overline{Se}$, $\overline{PPV}$, $\sigma_c$, $\delta_1$, $\delta_2$,
and $\rho$ are in line with the values used for the simulations.   

### Age rates
The median age rates ($r^I_i$, $r^P_i$, $Se_i$, and $PPV_i$) are
stored in the `rho` parameters, which are summarized as follows: 
```{r,eval=T}
  ## Age rates------
  ar_sum <- fit %>% gather_draws(rho[id_age,comp]) %>%
    group_by(id_age,comp) %>%
    select(-.variable) %>%
    tidybayes::summarise_draws("mean", "median","sd", "rhat","ess_bulk","ess_tail",
                               ~quantile(.x, probs = c(0.025, 0.975)))%>%
    rename(pm=mean,low=`2.5%`,up=`97.5%`) %>%
    ungroup() %>%
    mutate(comp=factor(comp,labels=c("r_i","r_p","se","ppv"))) %>%
    select(-contains("variable")) %>%
    decod_age() 
```

We can compare them with the true values used for the simulations,
showing the good agreement between the estimated rates and the true
rates used for the simulations:  

```{r,eval=T, fig.width=7, fig.height=7}
  ar_sum %<>%
    left_join(
      dt_sim %>%
        ungroup() %>%
        select(age,r_i=muI,r_p=muP,se,ppv) %>%
        pivot_longer(-age, names_to = "comp",values_to = "tm") %>%
        group_by(age,comp) %>%
        summarise(tm=median(tm))
    )
  
  ar_sum %>%
    qplot( data=., age, tm)+
    geom_line(aes(age,pm))+
    geom_ribbon(aes(ymin=low,ymax=up),alpha=.2)+
    facet_wrap(~comp, scales="free")+
    xlab("Age")+ylab("Rate")
  
  
```
  
### Spatial risks
The spatial risks $\theta_j^I$ and $\theta_j^P$ are stored in the
`theta` parameters, whereas the risks $\nu_1$ and $\nu_2$ for $Se$ and
$PPV$ are given in the `nu` parameters. 

```{r,eval=T}
  # ## spatial risks------
  spr_sum <- fit %>%
    gather_draws(nu[id_geo,comp],theta[id_geo,comp]) %>%
    tidybayes::summarise_draws("mean", "median","sd", "rhat","ess_bulk","ess_tail",
                               ~quantile(.x, probs = c(0.025, 0.975)))%>%
    rename(pm=mean,low=`2.5%`,up=`97.5%`) %>%
    ungroup() %>%
    mutate(comp=ifelse(is.na(comp),.variable,paste0(.variable,"_",comp))) %>%
    select(-contains("variable"))
  
  spr_sum %<>%
    group_by(comp) %>%
    mutate(n_geo=length(unique(id_geo))) %>%
    decod_geo()

```

Again, in our simulation context, we can compare the estimates with
the true values:

```{r,eval=T}
  tm_spr<-
    dt_sim %>%
    select(age,geo,RA,py,theta_1=muI,theta_2=muP,nu_1=se,nu_2=ppv,u1,u2) %>%
    group_by(age) %>%
    mutate(across(c(theta_1,theta_2),median,.names = "E_{.col}"),
           across(c(nu_1,nu_2),median,.names = "E_{.col}"),
           w=sum(py)) %>%
    group_by(geo,RA) %>%
    summarise( 
      se_b=sum(w*E_theta_1*E_nu_1)/sum(w*E_theta_1),
      ppv_b=sum(w*E_theta_2*E_nu_2)/sum(w*E_theta_2),
      nu_1 = sum(w*E_theta_1*nu_1)/sum(w*E_theta_1)/se_b,
      nu_2 = sum(w*E_theta_2*nu_2)/sum(w*E_theta_2)/ppv_b,
      theta_1 = sum(py*theta_1)/sum(py*E_theta_1),
      theta_2 = sum(py*theta_2)/sum(py*E_theta_2)
    ) %>%
    ungroup() %>%
    pivot_longer(-c(geo,RA), names_to = "comp",values_to = "tm") 
  
  spr_sum %<>%
    left_join(  tm_spr,
                by=c("geo","comp","RA"))
  
  spr_sum %>%
    group_by(comp,RA)%>%
    summarise(., r = sqrt(mean((pm - tm)^2)), sp = sd(pm), st = sd(tm),     
              sm = mean(sd), c = cor(pm, tm), p = 100 * (1 - r^2/st^2),     
              IP = 100 * mean((tm > low) * (tm < up)), 
              across(c(pm, tm), mean))

```

The alignment between the posterior means and the true values is
consistent with the simulations results presented in the manuscript; notably, the
95% credible intervals show satisfactory coverage. 

Finally, we can map the estimated risks; for instance, the spatial
distribution of incidence (areas covered by a registry are highlighted
in blue):

```{r,eval=T,fig.width=7, fig.height=4}
shap_ra <- shap_geo %>%
  filter(geo %in% (dt_sim %>% filter(RA)%$%geo %>% unique()))

theme_map <-
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        panel.spacing = unit(0,'lines'))

spr_sum %>%
  filter(comp=="theta_1") %>%
  select(geo,pm,tm) %>%
  pivot_longer(c(pm,tm),names_to = "typ") %>%
  left_join(shap_geo) %>%
  st_as_sf() %>%
  ggplot( data = ., aes(fill=value)) +
  geom_sf( colour = NA) +
    scale_fill_gradientn(colours=rev(RColorBrewer::brewer.pal( n = 9, name = "RdYlGn")),
                         na.value = "transparent",
                         limits=c(0.7,1.3),
                         oob = scales::squish)+
  facet_grid(~typ)+
  labs(fill=expression(theta))+
  geom_sf( data = shap_ra, colour = "blue",fill=NA) +
  theme_map+
  theme(legend.position="bottom")


```
